---
title: "玉米不同品种不同部位蛋白质含量分析"
subtitle: "汉玉122 vs 郑单958 | VE—V5 生育时期"
author: "数据分析报告"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-tools: true
    theme: flatly
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
---

## 引言

本报告分析两种玉米品种（**汉玉122** 和 **郑单958**）在营养生长期（VE—V5）不同器官的蛋白质含量差异及其动态变化。数据包含 5 个生育时期的 5 类器官（叶、茎、根、胚、胚乳）的蛋白质含量测定结果，每组 3 个生物学重复。

**研究问题：**

1. 不同器官之间的蛋白质含量是否存在显著差异？
2. 蛋白质含量在生育过程中如何变化？
3. 两个品种之间是否存在系统性差异？
4. 叶片内部的蛋白质空间分布规律是什么？
5. 不同茎节之间的蛋白质含量是否存在梯度？


# 数据读取与预处理

## 读取原始数据

```{r}
#| label: setup
#| code-summary: "加载R包与辅助函数"

library(tidyverse)
library(emmeans)
library(scales)
library(patchwork)
library(knitr)
library(kableExtra)
library(jsonlite)

# 器官配色
organ_colors <- c(
  "Leaf" = "#4DAF4A", "Stem" = "#377EB8", "Root" = "#E41A1C",
  "Embryo" = "#FF7F00", "Endosperm" = "#984EA3"
)

# 读取图片解读 JSON（首次渲染时文件不存在，跳过）
interpretations_file <- "data/figure_interpretations.json"
if (file.exists(interpretations_file)) {
  interpretations <- fromJSON(interpretations_file)
} else {
  interpretations <- NULL
}

# 显示解读结果的辅助函数
show_interpretation <- function(label) {
  if (!is.null(interpretations) && label %in% names(interpretations)) {
    cat(interpretations[[label]], "\n\n")
  }
}
```

```{r}
#| label: read-data

raw <- read_csv(
  "data/maize-protein-content.csv",
  col_types = cols(
    .default = col_double(),
    variety  = col_character(),
    part     = col_character(),
    stage    = col_character()
  )
)

raw <- raw |>
  select(-where(~ all(is.na(.)))) |>
  rename(replicate = relicate)

glimpse(raw)
```

## 数据清洗

```{r}
#| label: data-clean

df <- raw |>
  mutate(
    time = factor(time, levels = sort(unique(time))),
    stage = factor(stage, levels = c("VE", "V2", "V3", "V4", "V5")),
    variety_en = case_when(
      variety == "汉玉122" ~ "HY122",
      variety == "郑单958" ~ "ZD958",
      TRUE ~ variety
    ) |> factor(levels = c("HY122", "ZD958")),
    organ = case_when(
      part == "L"                 ~ "Leaf",
      part == "S"                 ~ "Stem",
      part == "R"                 ~ "Root",
      part == "Em"                ~ "Embryo",
      part == "En"                ~ "Endosperm",
      str_detect(part, "^L\\d")   ~ "Leaf",
      str_detect(part, "^S")      ~ "Stem",
      TRUE                        ~ NA_character_
    ),
    leaf_number = case_when(
      part == "L"                 ~ NA_real_,
      str_detect(part, "^L(\\d)") ~ str_extract(part, "(?<=^L)\\d") |> as.numeric(),
      TRUE                        ~ NA_real_
    ),
    distance = case_when(
      str_detect(part, "^L\\d([a-z])") ~ str_extract(part, "(?<=^L\\d)[a-z]"),
      str_detect(part, "^S([a-z])")    ~ str_extract(part, "(?<=^S)[a-z]"),
      TRUE                              ~ NA_character_
    ),
    stem_node = case_when(
      str_detect(part, "^S([a-z])") ~ str_extract(part, "(?<=^S)[a-z]"),
      TRUE                          ~ NA_character_
    )
  )

# 创建 figures 目录并设定导出路径
dir.create("figures", showWarnings = FALSE)
fig_dir <- "figures"
pptx_file <- "figures/protein_analysis.pptx"
```

## 数据概览

```{r}
#| label: data-summary

cat("观测数：", nrow(df), "\n")
cat("品种：", str_c(unique(df$variety), collapse = "、"), "\n")
cat("生育时期：", str_c(levels(df$stage), collapse = " → "), "\n")
cat("器官类型：", str_c(sort(unique(df$organ)), collapse = "、"), "\n")

df |>
  count(variety, stage, organ) |>
  pivot_wider(names_from = variety, values_from = n, values_fill = 0) |>
  kable(col.names = c("Stage", "Organ", "HY122", "ZD958"),
        caption = "各品种-时期-器官组合的样本数 / Sample counts by variety, stage, and organ") |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```


# 主分析：器官水平

## 器官间蛋白质含量差异

```{r}
#| label: fig-organ-boxplot
#| fig-cap: "**各器官蛋白质含量分布（全时期合并）** Distribution of protein content across five organ types for each variety, pooled over all growth stages. X-axis: organ type; Y-axis: protein content (mg/g); boxplots show median, IQR, and outliers; facets separate the two varieties."
#| fig-width: 10
#| fig-height: 5

p_organ <- ggplot(df, aes(x = organ, y = protein_content, fill = organ)) +
  geom_boxplot(alpha = 0.7, outlier.size = 1) +
  facet_wrap(~ variety_en, ncol = 2) +
  scale_fill_manual(values = organ_colors) +
  labs(x = "Organ", y = "Protein content (mg/g)", fill = "Organ",
       title = "Protein content by organ and variety") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, hjust = 1))

p_organ
ggsave(file.path(fig_dir, "fig-organ-boxplot.png"), p_organ, width = 10, height = 5, dpi = 300)
export::graph2ppt(p_organ, file = pptx_file, width = 10, height = 5, append = TRUE)
```

```{r}
#| label: result-organ-boxplot
#| results: asis
#| eval: true

show_interpretation("fig-organ-boxplot")
```

```{r}
#| label: anova-organ
#| code-summary: "器官与品种双因素方差分析"

aov_organ <- aov(protein_content ~ organ * variety, data = df)
summary(aov_organ)
```

```{r}
#| label: emmeans-organ
#| code-summary: "器官间两两比较"

emm_organ <- emmeans(aov_organ, ~ organ)
pairs(emm_organ, adjust = "tukey") |>
  as_tibble() |>
  mutate(
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE             ~ "ns"
    )
  ) |>
  kable(digits = 4,
        caption = "器官间蛋白质含量多重比较（Tukey HSD）/ Pairwise comparisons of protein content between organs") |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```


## 蛋白质含量随生育时期的变化

```{r}
#| label: fig-stage-trend
#| fig-cap: "**各器官蛋白质含量的生育时期变化趋势** Mean protein content (±SE) of each organ type across growth stages (VE to V5). X-axis: growth stage; Y-axis: protein content (mg/g); points connected by lines represent organ-specific trajectories; error bars represent standard error; facets separate the two varieties."
#| fig-width: 12
#| fig-height: 8

stage_means <- df |>
  group_by(variety_en, stage, organ) |>
  summarise(
    mean_se = mean(protein_content),
    se      = sd(protein_content) / sqrt(n()),
    .groups = "drop"
  )

p_stage <- ggplot(stage_means, aes(x = stage, y = mean_se, color = organ, group = organ)) +
  geom_point(size = 2.5) +
  geom_line(linewidth = 0.8) +
  geom_errorbar(aes(ymin = mean_se - se, ymax = mean_se + se),
                width = 0.15, linewidth = 0.5) +
  facet_wrap(~ variety_en, ncol = 1) +
  scale_color_manual(values = organ_colors) +
  labs(x = "Growth stage", y = "Protein content (mg/g, Mean ± SE)",
       color = "Organ",
       title = "Protein content trend across growth stages") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")

p_stage
ggsave(file.path(fig_dir, "fig-stage-trend.png"), p_stage, width = 12, height = 8, dpi = 300)
export::graph2ppt(p_stage, file = pptx_file, width = 12, height = 8, append = TRUE)
```

```{r}
#| label: result-stage-trend
#| results: asis
#| eval: true

show_interpretation("fig-stage-trend")
```

```{r}
#| label: anova-stage
#| code-summary: "三因素方差分析（organ × stage × variety）"

aov_full <- aov(protein_content ~ organ * stage * variety, data = df)
summary(aov_full)
```


## 品种间差异比较

```{r}
#| label: fig-variety-heatmap
#| fig-cap: "**品种间蛋白质含量差异热图** Heatmap of mean protein content difference between HY122 and ZD958 (HY122 minus ZD958) by organ and growth stage. X-axis: organ type; Y-axis: growth stage; color scale: blue (ZD958 higher) to red (HY122 higher); cell labels show numerical difference in mg/g."
#| fig-width: 10
#| fig-height: 5

variety_diff <- df |>
  group_by(stage, organ) |>
  summarise(
    mean_hy = mean(protein_content[variety == "汉玉122"]),
    mean_zd = mean(protein_content[variety == "郑单958"]),
    diff    = mean_hy - mean_zd,
    .groups = "drop"
  )

p_heatmap <- ggplot(variety_diff, aes(x = organ, y = stage, fill = diff)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.2f", diff)), size = 3.5) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    midpoint = 0, name = "Mean diff\n(mg/g)"
  ) +
  labs(x = "Organ", y = "Growth stage",
       title = "Protein content difference (HY122 − ZD958, mg/g)") +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

p_heatmap
ggsave(file.path(fig_dir, "fig-variety-heatmap.png"), p_heatmap, width = 10, height = 5, dpi = 300)
export::graph2ppt(p_heatmap, file = pptx_file, width = 10, height = 5, append = TRUE)
```

```{r}
#| label: result-variety-heatmap
#| results: asis
#| eval: true

show_interpretation("fig-variety-heatmap")
```

```{r}
#| label: fig-variety-paired
#| fig-cap: "**各生育时期两品种蛋白质含量对比** Side-by-side boxplots comparing protein content between HY122 and ZD958 for each organ, faceted by growth stage. X-axis: organ type; Y-axis: protein content (mg/g); red: HY122; blue: ZD958; boxplots show median, IQR, and outliers."
#| fig-width: 12
#| fig-height: 8

p_paired <- ggplot(df, aes(x = organ, y = protein_content, fill = variety_en)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(width = 0.8)) +
  facet_wrap(~ stage, ncol = 3) +
  scale_fill_manual(values = c("HY122" = "#E41A1C", "ZD958" = "#377EB8")) +
  labs(x = "Organ", y = "Protein content (mg/g)", fill = "Variety",
       title = "Protein content comparison by variety across stages") +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 40, hjust = 1))

p_paired
ggsave(file.path(fig_dir, "fig-variety-paired.png"), p_paired, width = 12, height = 8, dpi = 300)
export::graph2ppt(p_paired, file = pptx_file, width = 12, height = 8, append = TRUE)
```

```{r}
#| label: result-variety-paired
#| results: asis
#| eval: true

show_interpretation("fig-variety-paired")
```


# 细粒度分析：叶位与茎节（V2—V5）

以下分析聚焦 V2—V5 时期具有细粒度编号的叶片与茎节数据。

```{r}
#| label: subset-vg
#| code-summary: "筛选 V2-V5 细粒度数据"

df_leaf <- df |>
  filter(stage != "VE", organ == "Leaf", !is.na(leaf_number))

df_stem <- df |>
  filter(stage != "VE", organ == "Stem", !is.na(stem_node))
```


## 叶序与蛋白质含量

```{r}
#| label: fig-leaf-position
#| fig-cap: "**不同叶序的蛋白质含量分布** Protein content distribution by leaf number across growth stages. X-axis: leaf order (1 = lowest, higher = upper leaves); Y-axis: protein content (mg/g); red: HY122; blue: ZD958; facets separate growth stages; y-axis scales are free across facets."
#| fig-width: 12
#| fig-height: 8

p_leaf <- ggplot(df_leaf, aes(x = factor(leaf_number), y = protein_content,
                     fill = variety_en)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(width = 0.8)) +
  facet_wrap(~ stage, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = c("HY122" = "#E41A1C", "ZD958" = "#377EB8")) +
  labs(x = "Leaf order (bottom to top)", y = "Protein content (mg/g)", fill = "Variety",
       title = "Protein content by leaf position (V2-V5)") +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")

p_leaf
ggsave(file.path(fig_dir, "fig-leaf-position.png"), p_leaf, width = 12, height = 8, dpi = 300)
export::graph2ppt(p_leaf, file = pptx_file, width = 12, height = 8, append = TRUE)
```

```{r}
#| label: result-leaf-position
#| results: asis
#| eval: true

show_interpretation("fig-leaf-position")
```


## 叶内蛋白质空间梯度

```{r}
#| label: fig-leaf-gradient
#| fig-cap: "**叶片内离茎距离对蛋白质含量的影响** Intra-leaf protein spatial gradient by distance from stem. X-axis: distance from stem (a = nearest, f = farthest); Y-axis: protein content (mg/g); color-coded by distance; rows: growth stages; columns: variety and leaf number; y-axis scales are free."
#| fig-width: 12
#| fig-height: 10

distance_colors <- c("a" = "#1B7837", "b" = "#762A83",
                      "c" = "#E08214", "d" = "#2166AC",
                      "e" = "#B2182B", "f" = "#1A1A1A")

p_gradient <- df_leaf |>
  filter(!is.na(distance)) |>
  ggplot(aes(x = distance, y = protein_content, color = distance)) +
  geom_boxplot(alpha = 0.6, show.legend = FALSE) +
  facet_grid(stage ~ variety_en + factor(leaf_number), scales = "free") +
  scale_color_manual(values = distance_colors) +
  labs(x = "Distance from stem (a = near → f = far)", y = "Protein content (mg/g)",
       title = "Intra-leaf protein spatial gradient (by leaf and stage)") +
  theme_bw(base_size = 11)

p_gradient
ggsave(file.path(fig_dir, "fig-leaf-gradient.png"), p_gradient, width = 12, height = 10, dpi = 300)
export::graph2ppt(p_gradient, file = pptx_file, width = 12, height = 10, append = TRUE)
```

```{r}
#| label: result-leaf-gradient
#| results: asis
#| eval: true

show_interpretation("fig-leaf-gradient")
```


## 茎节差异

```{r}
#| label: fig-stem-node
#| fig-cap: "**不同茎节的蛋白质含量分布** Protein content by stem node position across growth stages. X-axis: stem node (a = basal, e = upper); Y-axis: protein content (mg/g); red: HY122; blue: ZD958; facets separate growth stages; y-axis scales are free."
#| fig-width: 10
#| fig-height: 8

p_stem <- ggplot(df_stem, aes(x = stem_node, y = protein_content, fill = variety_en)) +
  geom_boxplot(alpha = 0.7, position = position_dodge(width = 0.8)) +
  facet_wrap(~ stage, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = c("HY122" = "#E41A1C", "ZD958" = "#377EB8")) +
  labs(x = "Stem node (a = basal → e = upper)", y = "Protein content (mg/g)", fill = "Variety",
       title = "Protein content by stem node (V2-V5)") +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")

p_stem
ggsave(file.path(fig_dir, "fig-stem-node.png"), p_stem, width = 10, height = 8, dpi = 300)
export::graph2ppt(p_stem, file = pptx_file, width = 10, height = 8, append = TRUE)
```

```{r}
#| label: result-stem-node
#| results: asis
#| eval: true

show_interpretation("fig-stem-node")
```


# 补充分析

## 器官蛋白质分配模式

```{r}
#| label: fig-allocation
#| fig-cap: "**各品种器官蛋白质分配比例** Stacked proportion of mean protein content allocated to each organ type. X-axis: growth stage; Y-axis: proportion of total protein content (%); colors represent organ types; facets separate the two varieties."
#| fig-width: 10
#| fig-height: 6

alloc <- df |>
  group_by(variety_en, stage, organ) |>
  summarise(mean_p = mean(protein_content), .groups = "drop") |>
  group_by(variety_en, stage) |>
  mutate(prop = mean_p / sum(mean_p)) |>
  ungroup()

p_alloc <- ggplot(alloc, aes(x = stage, y = prop, fill = organ)) +
  geom_col(position = "stack", alpha = 0.85) +
  facet_wrap(~ variety_en, ncol = 2) +
  scale_fill_manual(values = organ_colors) +
  scale_y_continuous(labels = percent_format()) +
  labs(x = "Growth stage", y = "Protein content proportion", fill = "Organ",
       title = "Organ protein allocation across stages") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")

p_alloc
ggsave(file.path(fig_dir, "fig-allocation.png"), p_alloc, width = 10, height = 6, dpi = 300)
export::graph2ppt(p_alloc, file = pptx_file, width = 10, height = 6, append = TRUE)
```

```{r}
#| label: result-allocation
#| results: asis
#| eval: true

show_interpretation("fig-allocation")
```


## 胚与胚乳蛋白质含量动态

```{r}
#| label: fig-seed-protein
#| fig-cap: "**胚与胚乳蛋白质含量在营养生长期的变化** Embryo and endosperm protein content at VE and V2. X-axis: growth stage; Y-axis: protein content (mg/g); orange: Embryo; purple: Endosperm; circles: HY122; triangles: ZD958; diamonds: group means."
#| fig-width: 8
#| fig-height: 5

df_seed <- df |>
  filter(organ %in% c("Embryo", "Endosperm"))

p_seed <- ggplot(df_seed, aes(x = stage, y = protein_content,
                     color = organ, shape = variety_en)) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
  stat_summary(aes(group = interaction(organ, variety_en)),
               fun = mean, geom = "point", size = 4, shape = 18) +
  scale_color_manual(values = c("Embryo" = "#FF7F00", "Endosperm" = "#984EA3")) +
  scale_shape_manual(values = c("HY122" = 16, "ZD958" = 17)) +
  labs(x = "Growth stage", y = "Protein content (mg/g)", color = "Organ",
       title = "Embryo and endosperm protein content dynamics") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")

p_seed
ggsave(file.path(fig_dir, "fig-seed-protein.png"), p_seed, width = 8, height = 5, dpi = 300)
export::graph2ppt(p_seed, file = pptx_file, width = 8, height = 5, append = TRUE)
```

```{r}
#| label: result-seed-protein
#| results: asis
#| eval: true

show_interpretation("fig-seed-protein")
```


## 叶内蛋白质梯度强度量化

```{r}
#| label: gradient-summary
#| code-summary: "计算各叶片的近远梯度斜率"

gradient_data <- df_leaf |>
  filter(!is.na(distance)) |>
  mutate(dist_num = match(distance, letters[1:10])) |>
  group_by(variety, stage, leaf_number) |>
  summarise(
    slope = coef(lm(protein_content ~ dist_num))[2],
    r2    = summary(lm(protein_content ~ dist_num))$r.squared,
    n     = n(),
    .groups = "drop"
  )

gradient_data |>
  mutate(sig = ifelse(r2 > 0.3, "Strong", "Weak")) |>
  kable(digits = 3,
        caption = "叶片内蛋白质梯度 / Intra-leaf protein gradient (slope = change per unit distance)") |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```


## 蛋白质含量变异系数分析

```{r}
#| label: fig-cv
#| fig-cap: "**蛋白质含量变异系数** Coefficient of variation (CV%) of protein content by organ and variety across growth stages. X-axis: growth stage; Y-axis: CV (%); lines connect organ-specific CV values; facets separate the two varieties."
#| fig-width: 10
#| fig-height: 5

cv_data <- df |>
  group_by(variety_en, stage, organ) |>
  summarise(
    cv = sd(protein_content) / mean(protein_content) * 100,
    .groups = "drop"
  )

p_cv <- ggplot(cv_data, aes(x = stage, y = cv, color = organ, group = organ)) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~ variety_en, ncol = 2) +
  scale_color_manual(values = organ_colors) +
  labs(x = "Growth stage", y = "CV (%)", color = "Organ",
       title = "Coefficient of variation across stages") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")

p_cv
ggsave(file.path(fig_dir, "fig-cv.png"), p_cv, width = 10, height = 5, dpi = 300)
export::graph2ppt(p_cv, file = pptx_file, width = 10, height = 5, append = TRUE)
```

```{r}
#| label: result-cv
#| results: asis
#| eval: true

show_interpretation("fig-cv")
```


## 全株蛋白质积累趋势

```{r}
#| label: fig-total-protein
#| fig-cap: "**全株平均蛋白质含量变化** Whole-plant average protein content across growth stages for HY122 and ZD958. X-axis: growth stage; Y-axis: mean protein content (mg/g); error bars represent standard error; red: HY122; blue: ZD958."
#| fig-width: 8
#| fig-height: 5

total_means <- df |>
  group_by(variety_en, stage) |>
  summarise(
    total_mean = mean(protein_content),
    se         = sd(protein_content) / sqrt(n()),
    .groups = "drop"
  )

p_total <- ggplot(total_means, aes(x = stage, y = total_mean,
                         color = variety_en, group = variety_en)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  geom_errorbar(aes(ymin = total_mean - se, ymax = total_mean + se),
                width = 0.15, linewidth = 0.6) +
  scale_color_manual(values = c("HY122" = "#E41A1C", "ZD958" = "#377EB8")) +
  labs(x = "Growth stage", y = "Mean protein content (mg/g)", color = "Variety",
       title = "Whole-plant average protein content trend") +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")

p_total
ggsave(file.path(fig_dir, "fig-total-protein.png"), p_total, width = 8, height = 5, dpi = 300)
export::graph2ppt(p_total, file = pptx_file, width = 8, height = 5, append = TRUE)
```

```{r}
#| label: result-total-protein
#| results: asis
#| eval: true

show_interpretation("fig-total-protein")
```
